Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-11 output/paper/paper.tex
test

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode
% !BIB TS-program = biber
% !TeX root = ./paper.tex
\documentclass[a4paper,11pt,article,oneside,openany,american]{memoir}
\usepackage[memoir,bibliography]{setupscript}
\addbibresource{bibliography.bib}

% Customization
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{amsthm}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\begin{document}
\frontmatter%
\begin{titlingpage}
    \maketitle
    \begin{abstract}
        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}

	%\noindent Keywords:

\end{titlingpage}
%\tableofcontents*
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\mainmatter%

\chapter{Pandemic response demands timely data}
The new corona virus pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions (\cite{No_author_2020_coronavirus_three}). Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers (\cite{Jajosky2004_evaluation_reporting}). The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

@@

\section{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During a press conference, they present updates on the number of deaths, admissions to hosiptals and intensive care, as well as case counts.

@@

\chapter{The removal method}
We propose to use the removal method, developed in animal management (\cite{Pollock1991_review_papers}), to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s (\cite{Leslie1939_attempt_determine}). However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today (\cite{Matechou2016_open_models}). It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

@@

\chapter{Applying the model to COVID-19 in Sweden}
We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19. We then calculate the posterior distribution, prediction median and 95\% prediction intervals of the expected deaths from the reported deaths on each specific day. The method and algorithm is thoroughly described in the Supplementary Information.

@@

\section{Model Performance}
To say judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

@@

Benchmark and model point estimates are similarly close to the truth. The model produces tighter prediction intervals however, they are too tight between day 8-5 lags to reporting, see Figure \ref{fig:model_metrics}, this is likely due to that Swedish Public Health Agency does a database search during only certain days to match people reported dead to Covid infected. This type of reporting is currently not implemented in the model hence the poorer performance during these lags.

@@

\chapter{Implications and limitations}
The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.

@@

\chapter{Conclusion}
In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay. By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
\printbibliography%
\backmatter%
\appendix%
\chapter{Appendix}
\input{appendix.tex}%

>>>>

\documentclass[a4paper,11pth,american]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
\usepackage[babel]{csquotes}
\usepackage{babel}
\usepackage{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{hyperref}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\begin{document}
    \maketitle
    \begin{abstract}
        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The new corona virus pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions (\cite{No_author_2020_coronavirus_three}). Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers (\cite{Jajosky2004_evaluation_reporting}). The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

@@

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During a press conference, they present updates on the number of deaths, admissions to hosiptals and intensive care, as well as case counts.

@@

\section{The removal method}
We propose to use the removal method, developed in animal management (\cite{Pollock1991_review_papers}), to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s (\cite{Leslie1939_attempt_determine}). However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today (\cite{Matechou2016_open_models}). It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

@@

\section{Applying the model to COVID-19 in Sweden}
We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19. We then calculate the posterior distribution, prediction median and 95\% prediction intervals of the expected deaths from the reported deaths on each specific day. The method and algorithm is thoroughly described in the Supplementary Information.

@@

\subsection{Model Performance}
To judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it is biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

@@

Benchmark and model point estimates are similarly close to the truth. The model produces tighter prediction intervals. For 8-5 days of reporting lag (see Figure \ref{fig:model_metrics}), the intervals are too tight. This is likely because the Public Health Agency queries the Swedish death registry for Covid-19 deaths only once or twice a week. Since we do not know the process, it has not been explicitly modeled.

@@

\section{Implications and limitations}
The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.

@@

\section{Conclusion}
In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay. By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
\bibliographystyle{chicago}
\bibliography{bibliography}
\appendix
\section{Appendix}
\input{appendix.tex}%
